---
title: "Day3_DataTypes_DataStructures"
pagetitle: "Data Types and Structures in R"
output:
  html_document:
    code_folding: show # allows toggling of showing and hiding code. Remove if not using code.
    code_download: true # allows the user to download the source .Rmd file. Remove if not using code.
    includes:
      after_body: footer.html # include a custom footer.
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
date: "2025-02-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Data Organization in R

This section covers how data is organized in R. This includes both types of data (i.e., is your data composed of numbers, words, or both) and data structures (i.e., do you have one long list of elements, a spreadsheet, or something else). This section also covers how to recode existing data into another type of data and how to deal with missing data. Let's get started! 

## Data Types in R
Data can come in many different formats or units. Sometimes we have numbers with decimals, sometimes we have counts of how many times something happened, sometimes we have categories with names. It is important to know what type of data you have, how to figure that out in R, and how to modify the data type if needed. Various functions in R require having certain data types, so this is an important foundational skill for managing data in R. 

#### Numeric
Numeric data is data made up of numbers. This is what we most commonly think of as data. For example, how many hours people slept on a given day (6, 8.25, 7.5), the percent accuracy that students got on an exam (87.45, 75.92, 98.40), or number of people who attended a data management workshop each day (32, 38, 28). 

#### Integer
This data type deals with whole numbers (i.e., no decimals) and operates similarly to numeric data types. Some data is stored as "numeric" and "integer" at the same time.

#### Character/String
Character variables are variables with words or letters. Such as names of people ("Erin", "Maria", "Huy") or places ("ON", "BC", "QC). Character variables usually need to be put in quotation marks in R code.

#### Boolean/Logical
Boolean and logical operators can be used to compare values or create logical statements. You might have used Boolean characters to search databases in the past (e.g., searching in journal databases for articles with certain keywords). They work similarly in R. These functions also can be useful if you want to filter for a certain value in the data (e.g., sub-set only women in a dataset) or modify how data is grouped (e.g., you want to re-label everyone over 6 feet as "tall").

Symbol | Function
|-------|---------|
| > | Greater than|
|< | Less than|
|>= | Greater than or equal to|
|<= | Less than or equal to|
|== | Equal to|
|!= | Not equal to|
|& | AND|
|! | NOT|
|TRUE | A statement is true|
|FALSE | A statement is false| 

## Missing Data
Some missing data is usually present in data. There are many reasons for missing data and many different ways to handle missing data. These choices will depend on the context of specific research, but generally a plan for how to approach missing data should be thought through in advance. 

In this dataset missing data has been coded in a variety of ways, capturing the reason for the data being missing. For example, for several of the categorical variables, such as age group, if the participant refused to answer the question the data includes the code "98," while if the question was skipped for a valid reason the code "96" was applied. 

For questions about duration, where the variables are in minutes (e.g., minutes they have slept), if the participant refused to answer the code "9998" was used, or if the participant said they didn't know the code "9997" was used. 

You will notice that in both cases, the number used to show that this data is missing is well out of the range of the actual complete data (i.e., the age groups are coded 1 through 7, and people would have to sleep for a whole week to reaching that many minutes). 

However, we may run into an issue if we start to do analyses and include these numerical codes as actual points of data (i.e., find the mean of duration of sleep). To avoid this, we will recode these values that indicate missing values to read as "NA" instead. 

```{r}
# data <- data %>% mutate (across (
#   
# ))
# durSleep, 
# durMealPrep, 
# durEating, 
# durAlone, 
# durDriving, 
# durWork, 
# durSchoolSite, 
# durSchoolOnline, 
# durStudy,
# )

#> 9995 means missing data
# 9996 = Valid skip, 9997 = Don't know, 9998 = refusal, 9999 = not stated
# 
# mutate(across(
#         y1:y2,
#         ~na_if(., 999)
#     ))
```
### Useful Missing Data Related Functions
If your data contains NA values, sometimes functions will give error messages due to incomplete data. Including `na.rm` to the function will typically tell R that it should complete the function, ignoring missing values. 

If you want to check if a varibale or data frame has missing data, you can use the function `is.na`. 

If you want to find instances where there is no missing data you can use `complete.cases()`. 

na.rm, is.na, complete.cases()
read in as NA? - yes, in Day 3 part 1, cover that in the integer → character variables + the duration variables (teach once, copy/paste code for the rest), check work via the commands above 

## Introduction to dplyr Library
dplyr is a powerful R package designed to simplify data manipulation tasks. It provides a set of intuitive functions that make filtering, selecting, and transforming data frames (we will talk about this data structure soon) more readable and efficient. It is the go to package for data set organization. 

Lets load the dplyr library.

```{r, warning = FALSE}
library(dplyr)
```
Here are some of the most common functions of dplyr. We will practice using some of these today with our dataset. 

  - `mutate()`: Useful when you want to create new columns based on values in existing columns in your dataset. For example, you may want to recode numeric data into new categories with character names. 
  - `recode()`: Useful for changing data format of variables or re-categorizing certain values in a variable as some other value. For example, maybe you want to label everyone who sleeps less than seven hours a night as "sleep deprived." 
  - `group_by()`: Allows you to view your data based on groups that you define or already exist in your data. For example, you may want to group your data based on whether people live in a city or rural area, or by gender. 
  - `summarize()` or `summarise()`: Useful for seeing basic descriptive info about your data. This is particularly helpful when use with the `group_by()` function. For example, you can see if people in cities compared to rural areas feel more rushed on average (i.e., compare means between these two groups). 
  - `%>%` (pipe operator): Allows you to tell R to apply one function several times. You can read it more or less as "and then" when thinking about what your R code is doing. 
  - `filter()`: This function allows you to extract rows that meet specific conditions you include. 
  - `select()`: This function lets you choose particular columns based on your specifications. 
  
It is important when using dplyr to assign your output to an object if you want to retain the info you've found or created. Remember the `<-` symbol can be used to assign output to a named object. We will practice this today too. 

Together, these functions in dplyr enable you to easily explore your data by various subgroups, create new variables, or create a focused subset of your data for further analysis.

Let's explore the data types in our data using the `dplyr` functions! 

## Checking Data Types in Our Data

First we load our data. Information on the data type can be seen in the "head" function in R. NEED TO GET THIS TO PRINT ON WEBSITE.
```{r}
load("data/timeuse_day2.Rdata")
data <- mydata
print(head(data))
```

You can also check the data type of a specific variable with the "class" function,

```{r}
print(class(data$province))
```

## Recoding Variables and Changing Data Types
For some variables, we might want to change the data type based on our codebook. If we check the current data type of our "province" variable we see that it is in integer format. 
```{r}
print(class(data$province))
```

However, maybe we would prefer to have province acronyms (i.e., character data format) rather than the integer representation of them. We can re-code variables using the dplyr functions `%>%` and `mutate()`.

This code is telling R to `recode()` the variable "province" from the dataset "data" into a new variable called "province_char" (you could call it anything, but make sure it is a meaningful name). The list outlines how each value of the current "province" variable should be recoded for the new "province_char" variable. These integer values and what province they represent can be found in the dataset codebook. The `mutate()` portion tells R to create a new column called "province_char". Before the `mutate()` function this code tells R to take the "data" object (i.e., dataset) "and then" (i.e., the pipe function) apply the following functions. The pipe function (`%>%`) tells R to recode all of these values all at once, rather than having to write out `mutate()` and `recode()` for each one. Lastly, all of this work needs to be assigned back to our dataset "data" in order to save it. Note that we did not assign this work to replace/overwrite the existing "province" variable. In order to have transparent data management and to prevent overwriting the original data in the instance that you make a mistake in your recoding, it is always best practice to keep the original variables in their original format and create new variables based on them as needed. In this way, if you ever need to reference the original data (and you definitely will need to), you still have it available to you. 
```{r}
data<-data %>% mutate(province_char = recode(data$province, 
                            "10" = "NL", 
                            "11" = "PEI", 
                            "12" = "NS",
                            "13" = "NB", 
                            "24" = "QC", 
                            "35" = "ON", 
                            "46" = "MB", 
                            "47" = "SK", 
                            "48" = "AB", 
                            "59" = "BC", 
                            "96" = "NA", 
                            "97" = "NA", 
                            "98" = "NA", 
                            "99" = "NA"))
```

Now lets check the data type of this new variable we just created. 
```{r}
print(class(data$province_char))
```

Most of this dataset is in integer format, because the data does not have decimals, and that is what R defaults to. However, it also stores this data as numeric at the same time. You can see this by checking the "class" of the data and asking if the data "is.numeric".

```{r}
print(class(data$durSleep))
is.numeric(data$durSleep)

```
You will see here that you get a logical operator as the output. 

Sometimes you may want to change the data type without recoding all of the data. In this case a useful function is `as.numeric()`. Because this data is already stored as both integer and numeric, this function will not change the data type in this instance, but this can be helpful for other datasets you may work with in the future. Here is an example of how you would use this function. 
```{r, include = FALSE}
as.numeric(data$durSleep)
```

## Data Structures in R
Data structures are how the data is organized in R. Here are the major data structures in R. Some of these are mostly used only when doing statistical modeling, but we will list them all out here anyway. 

#### Vectors
Vectors are simplest data structure. They are essentially a list of elements of the same data type. For example, you could have a numeric vector containing the elements (10, 15, 20, 25) or a character vector containing the elements ("blue", "red", "green"). They are one dimensional, since they are just one long list. A single variable (e.g., a list of ages of all the participants) can be considered a vector. 

#### Matrices
Matrices, like vectors, can contain only a single type of data. However, matrices are two-dimensional (i.e., they have rows and columns). This is a data structure R uses to store certain information about statistical models. A matrix of numbers could look like this: 

```{r}
matrix_example <- matrix(
  c(1, 2, 3, 4, 5, 6, 7, 8, 9), 
  nrow = 3,             
  ncol = 3,             
  byrow = TRUE)
print (matrix_example)
```

#### Arrays
Arrays are an extension of matrices, in that they can have additional dimensions. For example, an array that is 3:4:2, has three rows, four columns, and two layers in depth. You likely won't need to use arrays much, but know that when conducting statistics sometimes, the model results are stored in matrices and arrays. 

#### Lists
Lists are an ordered collection of elements. They can contain different data types and they are one dimensional. So, you can have a list of vectors, a list of characters etc. Here is an example of a list:

```{r}
list_example <- list("red", "blue", "skyblue", "forestgreen")
print(list_example)
```

Because lists are ordered (i.e., each element in the list has a specific spot in the list), you can call for an element based on the order it has been assigned. For example, "skyblue" is third in this example list. 

```{r}
print(list_example [3])
```

#### Data Frames
Data frames are the most common and popular type of data structure used in R. Data frames work essentially like spreadsheets in that they are two dimensional (i.e., have columns and rows) and can store different data types in teh same data frame (i.e., numeric and character). 

However, data frames do have some constraints. First, the columns and rows must be named. Second, the lists of vectors in the data frame must have equal lengths (i.e., each column needs an identical number of elements/rows). Third, all elements in a single column must be of the same data type (i.e., you can't switch between numbers and names in a single column)

For this course, our data is in data frame format. 

#### Factors
Factor data is good for categorical/string data. Factors can store string or integer data. Factors have leveled data. For example, the variable of education may have the levels of "high school," "bachelors degree," and "masters degree." 

Factor structure is particularly useful for statistical modeling. Sometimes you might need to change an object into factor structure, so you can use the function `as.factor()` to do so. 

#### Tibbles
Tibbles are an updated version of data frames and part of the tidyverse package. They are a bit more user-friendly version of tabular data. 

## Your Turn!
### Changing Data Types
::: question
What type of data is the variable for marital status?
:::

```{r class.source = 'fold-hide', results = 'hide'}
typeof(data$maritalStat)
```

:::question
Can you re-code marital status into a character/string variable?
:::

```{r class.source = 'fold-hide', results = 'hide'}
data<-data %>% mutate(maritalStat_char = recode(data$maritalStat, 
                            "1" = "Married", 
                            "2" = "Living common-law", 
                            "3" = "Widowed",
                            "4" = "Separated", 
                            "5" = "Divorced", 
                            "6" = "Single, never married", 
                            "96" = "NA",
                            "97" ="NA", 
                            "98" = "NAl",
                            "99" = "NA"))

class(data$maritalStat_char)
```

::: question
How about level of education?
:::

```{r class.source = 'fold-hide', results = 'hide'}
data <- data %>% mutate(eduLevel_char = recode(data$eduLevel,
                            "1" = "Less than high school dimploma or its equivalent",
                            "2" = "High school diploma or equivalency",
                            "3" = "Trade certificate or diploma",
                            "4" = "College, CEGEP, or other non-university certificate or dimploma",
                            "5" = "University certificate or dimploma below the bachelor's level",
                            "6" = "Bachelor's degree",
                            "7" = "University certificate, diploma, degree above the BA level",
                            "96" = "NA",
                            "97" = "NA",
                            "98" = "NA", 
                            "99" = "NA"))

class(data$eduLevel_char)
```
### Check Your Knowledge of Data Structures

:::question
What data structure would you use to store the following data?
a) A list of heights of movie characters

b) A list of names of characters from a movie

c) A spreadsheet with both a and b in it

d) The data we have been working with in this workshop
:::

:::question
Is our data in a data frame format? Hint: use the function "is.data.frame()"
:::

