---
title: "js_timeuse_20250514"
output: html_document
date: "2025-05-14"
---

```{r}
#Define the directory that has the dataset from previous activity
file_path <- "data/timeuse_day3_1.Rdata"

#Load file into R object
load(file_path)

#Assign this dataframe to object name js_data
js_data <- data

#See the first few rows of js_data
js_data |> 
  head(10)
```

Let's view all the column names. 

```{r}
colnames(js_data)
```
Question: How do people who feel rushed spend their time different from those who don't 

Columns that we're interested in: `feelRush`, all others that start with `dur` (e.g., `durSleep`, `durAlone`, `durWork`)

## Take a look at the feelRushed column. 

```{r}
library(dplyr)
```


```{r}
# Distinct function of dplyr will print out the number of unique values of a particular columns
js_data |> 
  distinct(feelRushed)

# Count function of dplyr will print out the number of unique values of particular columns with their counts

js_data |> 
  count(feelRushed)
```


## Let's take a look at dur columns

```{r}
js_data |> 
  count(durSleep) |> 
  arrange(desc(n))
```

```{r}
js_data |> 
  count(durWork) |> 
  arrange(desc(n))
```

| Code | Value            |
|------|------------------|
| 1    | daily            |
| 2    | few Times a Week |
| 3    | once a Week      |
| 4    | once a Month     |
| 5    | less a Month     |
| 6    | never            |


Feel Rushed: more than once a week (e.g. feelRushed <= 3)
Do not feel Rushed: less than once week (e.g., feelRushed >= 3)

```{r}
# Filter all rows of feelRushed column that are <= 3

rushed <- js_data |> 
  filter(feelRushed <= 3)

not_rushed <- js_data |> 
  filter(feelRushed > 3)

nrow(rushed)
```

Aim: Filter the rows that have durSleep are between 300 and 600
```{r}
# between 300 and 600: >= 300 and <=600 
sleep_range <- js_data |> 
  filter(durSleep >= 300 & durSleep <= 600)

sleep_range

# All the rows with durSleep is either smaller than 300 or larger than 600
sleep_range <- js_data |> 
  filter(durSleep < 300 | durSleep > 600)

sleep_range

```
Feel Rushed: more than once a week (e.g. feelRushed <= 3, or 1, 2, 3)
Do not feel Rushed: less than once week (e.g., feelRushed >= 3 or 4, 5, 6)

Feel Rushed are those who has feelRushed either is equal to 1, or 2, or 3

```{r}
rushed <- js_data |> 
  filter(feelRushed == 1 | feelRushed == 2 | feelRushed == 3)

nrow(rushed)

```


```{r}
rushed_time <- rushed |> 
  select(durSleep, durAlone, durWork)

# all column that start with "dur": selections hepers: 

not_rushed_time <- not_rushed |> 
  select(durSleep, durAlone, durWork)
```

# Very simple analyses: compare the mean of these columns between rushed and not rushed group 


```{r}
avg_rushed_time <- rushed_time |> 
  summarise(durSleep=mean(durSleep),durAlone=mean(durAlone),durWork=mean(durWork))

avg_not_rushed_time <- not_rushed_time |> 
  summarise(durSleep=mean(durSleep),durAlone=mean(durAlone),durWork=mean(durWork))

rushed_time_all <- rushed |>
  select(starts_with('dur'))

avg_rushed_time_all <- rushed_time_all |> 
  summarise(across(everything(),mean))
```


```{r}
diff <- avg_rushed_time_all - avg_not_rushed_time_all
```

```{r}
diff
```



